---
title: "Introduction to R and RStudio- Workshop 2 "
author: "Aybuke Atalay & Alex Crest"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## Intro to R and R Studio: Workshop 2

Welcome back! I hope you are all ready to continue on our R journey.
Today, we are looking to build on our foundations from last week, and to
provide you with the tools to use R for conducting and communicating
your own research.

We will be working with the amazing
[PalmerPenguins](https://allisonhorst.github.io/palmerpenguins/) dataset
today. This data was collected by Allison Horst and their team, and
investigated the physiological attributes of 3 penguin species from 3
different islands.

### Learning aims

We will work through the research questions to guide us through our own
learning aims for today:

-   Setting projects and managing our data/documents.
-   Uploading data.
-   Getting to know the tidyverse package.
-   Making the pipe (%>%) your best friend.
-   Tidy and filter our data using R.
-   Summarising data with R.
-   Saving data as a .csv file


## Project setting

For this stage, we will be temporarily abandoning our document as the
new project is set. But don't worry, we will shortly reopen this
document within said project. We will need to follow these steps:

1)  Go to the `File` tab, and select `New project...`.

2)  Select `New Directory`, and from there select `New Project`.

3)  This will take us a `Create New Project` pane. Choose the folder you
    want the project to live in, and name your project.

4)  Our new project is made! At this point, we may want to reopen this
    document, so we can follow the steps again.

5)  Now we want to pay attention to our Files/Plots/Packages/... Panel
    (Usually bottom right of the R studio screen). Click the `Files`
    tab, and click the `+NewFolder` button to make some new folders.

6)  We want to make the following folders:

    -   raw_data
    -   prepared_data
    -   Documents
    -   Plots
    -   Other (for any kind of other file or output you want to store).

7)  And there we go. We are ready!


## Traditional setup
We will be using the following packages for today's workshop
```{r}
install.packages("palmerpenguins")
install.packages("tidyverse")
install.packages("gt")
```


### Checking Installed Packages
Before running the code above, you can check if you already installed those packages!

```{r}
print(requireNamespace("palmerpenguins", quietly = TRUE))
print(requireNamespace("tidyverse", quietly = TRUE))
print(requireNamespace("gt", quietly = TRUE))
```

### Load Packages
```{r}
library(palmerpenguins) # Contains the Palmer Penguin dataset.
library(tidyverse) # Contains multiple very useful packages for data tidying and visualisation. Essential for almost any project.
library(gt) # Make pretty tables
```


## Uploading data

We have our research questions. And we have our learning aims. Lets crack on with the important first step of data analysis: uploading our data!

The Penguins dataset is loaded to your workspace when you run the "library(palmerpenguins)" command. This makes the dataset nice and user friendly to get our head around playing with data. 
To load the data into your environment, you have multiple options:

Option 1 (Works with .csv documents, slightly quicker):
Go to `Files`
Click on the penguin.csv document at `Files` then click on`Import Dataset`
Click on `Import`

Option 2 (Works with multiple types of documents):
Find the `Environment` panel, and click on the `Import Dataset` tab.
From there, click `From Text (readr)...`. (We can also upload excel,SPSS, SAS and STATA data files from here as well).

Option 3:
```{r data_upload, echo = FALSE}
library(readr)
penguins <- read_csv("penguins.csv")  # We will copy the path directory from the Import Dataset tab, and paste it here.
```
Option 4
```{r}
penguins <- read_csv(url("https://raw.github.com/DCS-training/IntroToRAndRStudio/main/2024/Data/penguins.csv")) # This is our back-up option. It will take the dataset directly from github, allowing you to follow the materials if there is an issue with importing the data

```

Congratulations! You have uploaded your dataset into R. Now it's time to prepare our data for analysis.


## Data Tidying in R

Data tidying has a bad rep. It is perceived as long and tedious. It's also a space where you have to carefully track your steps, so you can ensure that your analysis is reproducible and reliable. This is an area where R thrives.

Using the `Tidyverse` functions, we have a variety of very useful tools to ease up our tidying workflow. The function names are descriptive, and the ability to comment with `#` in the code chunks and write notes in the markdown document makes our tidying script a logbook of our steps.

My top tip for any of you starting your data collection is to use the early stages as an opportunity to build and develop your R skills through data tidying. It may feel like a steep learning curve to begin with, but every step you work through now will save you mountains of time in the future.

Your raw data will remain raw, and so you don't need to worry about accidentally vandalising it. And the ability to save your data means you can instantly use it in any way you like in the future - even if that means using Excel or SPSS to do a step you can't do in R (we won't judge too much...).

It is well worth the investment of time, especially if you are in the early stages of data collection. There is plenty of help online, and you can always get in touch with [CDCS through the data surgery service](https://www.cdcs.ed.ac.uk/training/data-surgery) if you are particularly stuck.

### Inspecting our data

Before tidying our data, we need to know our data. Let's use R to help us know our data. We will do in this two stages. First with the data itself. And second through visual methods.

Things to look out for:

\* Presence of NA's

\* Strange values

\* Checking if the data is correctly structured

Let's inspect our dataset!


#### numeric data inspecting

```{r}
View(penguins) # Lets us view the entire dataset. Not advised for large data.
head(penguins) # Lets us view the first 6 rows
tail(penguins, n = 8) # Lets us view the last 6 rows
summary(penguins) # Provides an overview of all our variables
glimpse(penguins) # Get an overview of the data structure along with a preview of the first few rows in each variable.

```



### Inspect Rows and Columns at the Data
```{r}
nrow(penguins) # prints the number of rows 
ncol(penguins) # prints the number of columns 
names(penguins) # prints the column names
```

### Inspect Variable Types
```{r}
class(penguins$species)
class(penguins$body_mass_g)
class(penguins$year)
```


#### Visual Data Inspection

```{r}
hist(penguins$flipper_length_mm) # using $ allows us to choose our variable. Try Changing flipper_length_mm to bill_length_mm, or any other variable from the dataset!
boxplot(penguins$flipper_length_mm)
```

## Task time 1

-   Using `summary()` and `str()` identify variables with NA
-   Identify if all variables are correctly structured.
-   Use `hist()` and `boxplot()` to check the distribution of
    `body_mass_g`.

```{r}
summary(penguins)
glimpse(penguins)
hist(penguins$body_mass_g)
boxplot(penguins$body_mass_g)
```



## Preprocessing our data

We have now identified our missing data, and inspected our variables for outliers and their distributions. It's time to process our data to get it ready for analysis!

Here we will make use of the `dplyr` tools from the `tidyverse` family of packages. You can learn more about the useful tools
[here](https://dplyr.tidyverse.org/articles/dplyr.html).

We will also make use of the `%>%` pipe that we introduced last week. This makes it very easy to combine multiple functions, as it takes the object to the left, and applies the next function to it.

Let's put this into action so you can see what happens!

### Identifying and Cleaning Missing Variables :  is.na() and drop_na()

```{r}
sum(is.na(penguins)) # Total number of Nas in the document
sum(!complete.cases(penguins)) # total number of rows that contain at least one NA
penguins_clean<- penguins %>% drop_na()

```


### Creating new variables: mutate()

Mutate is used to change our variables. Think of it as "Transform variables" from SPSS, but more versatile! We can use a variety of functions within mutate, and even create our own when we get advanced. We can also combine functions, and use newly created variables in new commands. It's such a powerful tool. 

It's also very useful for any researchers working with psychometric measures - as it allows us to create total scores whilst keeping track of our steps within the code.

```{r}
mutate_example <- penguins_clean %>%
  mutate(
    body_mass_kg = body_mass_g/1000, # here we use mutate to convert the units from grams to kilograms 
    species = as.factor(species), # We can use R to convert our data from character to factor format
    bill.length.mm = bill_length_mm, # we can also use mutate to create a new variable with a different variable name
    flipper_length_Z = scale(flipper_length_mm)  # standardise our data
  )
```

### Creating a new variable with condition using mutate(): `if_else` and `case_when` functions 

```{r}
mutate_example_2 <- penguins_clean %>%
  mutate(year_2009 = if_else(year==2009, "Yes", "No"))
  # if year=2009, the new column (year_2009) records as "Yes"; if the year != 2009, the new column (year_2009) records as "No"
  

mutate_example_3 <- penguins_clean %>%
                      mutate(flipper_length_group = case_when(flipper_length_mm > 200 ~ "Big Penguinz",
                                           flipper_length_mm >= 180 & flipper_length_mm <= 200~ "Mid Size Penguinz",
                                           flipper_length_mm < 180 ~ "Small Penguinz"))
```



### Renaming a Column Name: rename()

We can also rename our variables using the `rename()` function. This is useful if we want to replace our existing variable names (whilst using mutate() will keep a copy of the old name within the data) 

```{r}
renamed_varaibles <- penguins_clean %>% 
  rename("bill.length.mm" = bill_length_mm)
```

### Filtering rows: filter()

There are times where we need to filter our data. We may want to focus on certain populations. Or we might want remove outliers/suspicious values. All this can be done with `filter()`. 

Filter operates by logic rules. Here are some examples:

* `==` selects data that matches the exact criteria
* `!=` removes data that matches the following criteria
* `<=` select values equal to or less than our criteria
* `>=` select values equal to or greater than our criteria

We can also combine filter rules in R. We can specify *AND* with either `&` or `,`. Meanwhile, we specify *OR* with `|`. 

```{r}
## Filter examples

Adelie_Penguins <- penguins_clean %>%
  filter(
    species == "Adelie"
    )

Big_Gentoo_Penguins <- penguins_clean %>%
  filter(
   species == "Gentoo",
   body_mass_g > 5000
  )

Biscoe_OR_Dream_island <- penguins_clean %>% 
  filter(
    island == "Biscoe" | island == "Dream"
  )


```

### Selecting Columns: select()
```{r}
penguin_islands<- penguins_clean %>%
                  select(species,island)
head(penguin_islands)

```


## 10 min Break Time

Grab yourself a coffee, and have a chill. Feel free to ask us any questions during this time!


### Putting it all together using %>%

Whilst we can work iteratively to change our data step by step, we can also combine multiple functions with the pipe `%>%`. 

```{r}
tidy_data <- penguins_clean %>% 
  mutate(  
    year = as.factor(year), 
    species = as.factor(species),
    island = as.factor(island),
    body_mass_kg = body_mass_g/1000
    ) %>%
  filter(species == "Adelie") %>% 
  rename("row_id" = `...1`)  # renaming the weird first column of row numbers, the syntax for `rename()` is new_name = old_name
 
  
summary(tidy_data) # always produce a summary of your data after tidying - check if at all makes sense.
```

### Example with arrange() and slice()
```{r}
top_10_biggest <- penguins_clean %>%
                arrange(desc(body_mass_g)) %>%
                slice(0:10)

top_10_smallest <- penguins_clean %>%
                arrange((body_mass_g)) %>%
                slice(0:10)

```

### Grouping and Summarising Data: group_by() and summarise()

```{r}
penguins_summary<- penguins_clean %>% 
                    group_by(species) %>%
                    summarise(n=n(), # number of observations in each species
                              mean_weight = mean(body_mass_g)) # average weight of species


# Creating a table 
penguins_summary %>% 
  gt()
  
```


## Saving our data

Now that the data is processed, we need to save it. This allows us to
save time in future steps, and provides us with the freedom to use other
software to work with your data if needed (again, we won't judge...).

Saving our data is easy, but there are some very important steps to
take: 

1) Making sure we have our working directory set. This is where
our data will be saved to.
2) Protect your raw data. Make sure your new file name does not overwrite your raw data. 
3) I always comment out my code after saving. This extra step helps to prevent accidental overwriting.

Time to run our code and save our data.

```{r}
write.csv(tidy_data, file= "prepared_data/tidy_data.csv") # the object you want to save and the path of the file you want to save into
```


### Task 2

Using penguins_clean

- Create a new variable called `bill_surface_area_mm2` by multiplying `bill_length_mm` with `bill_depth_mm`. (Hint - in R, we use * to multiply variables).
- Filter `species` and only keep "Adelie" and "Gentoo"
- Compare the average `bill_surface_area_mm2`. Tip: use group_by() and summarise() functions from dplyr
- Visualise the distribution of `bill_surface_area_mm2` using hist()
- Interpret the visualisations. What do we think?
- Save the new data you created to prepared_data folder!



### Bonus Materials

### Data Visualisation

We appreciate that we are likley to run out of time during this course, as there are so many exciting things to cover. One of them is data visualising (as showcases above). If you want to learn in more depth about using R for data visualisation, please check our [CDCS resources on Good Data Visualisation with R](https://github.com/DCS-training/Good-Data-Visualisation-with-R). 

After preprocessing data,you can try to visualise our analysis to help communicate your data and some summary statistics
with greater clarity using `ggplot` 

We also recommend playing with this amazing [online gui ggplot
app](https://shiny.gmw.rug.nl/ggplotgui/), as it gives you a graphic
interface to create your plots, and provides you with the code to
generate it in R!

Below is an example of a ggplot output. 

```{r}
ggplot(data = tidy_data, # select data
       aes(x = sex, y= body_mass_kg, fill = sex)) + #select your variables
  geom_boxplot(width = .3, alpha = .7 ) + # Setting a boxplot to visualise data
  theme_minimal() + # I like the minimal theme, as it declutters the plot, and meets the APA styling requirements
  theme(
        legend.position = "none" # removing legend. Try replacing "none" with "bottom"
        ) + 
  labs(title = "Does Penguin Bodymass differ by sex?",
       subtitle = "Male Penguins (M = 4.55, sd = .79) are significantly heavier ( t = 8.55, df = 323.9, p <.05).
than female penguins (M = 3.86, sd = .67)",
       x = "Sex",
       y = "Body Mass (kg)",
       caption = "Data from PalmerPenguins"
  )
```


Below, we have materials below to take you step by step through the layers of creating plots with R. It has so much potential for customising, and allows for near infinite versatility in data visualisation. I love it, and will never go back to excel for my plots. We hope you enjoy! 

## Plotting with ggplot

By now, we have seen some of the plotting we can do with R. Now it's
time to take you through the process 1 step at a time, so that you can
have the confidence to play with R and make your own wonderful visual
masterpieces.

### Step 1 : Assign your data

Here we need to tell R which dataset we want to use. We will load this
data into the ggplot function.

```{r}

ggplot(data = tidy_data)

```

### Step 2: Assigning variables

Now the data is loaded, it's time to assign our variables. These
variables are all assigned inside the `aes()` of the ggplot command. We
have so many options to play with in R. Here are some of available
choices we can use:

-   x
-   y
-   color/colour
-   fill
-   size
-   shape
-   linetype

Remember - just because there is an option to use, doesn't mean we
should use it... Except for today. Today, play with everything to see
what happens.

```{r}

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       )

```

### Step 3 - Assigning our geometry

It's all well and good setting our variables. But until we give R some instructions on how to present it, it can't do anything. We provide these instructions using geoms (for geometry - geometry and statistics have many overlapping mathematical features, which makes it perfect for visualising analyses).

Our geom options include, but are not limited to: 

- `geom_point()`: apply a scatter chart 
- `geom_smooth()`: apply predictive model (i.e., line of best fit) 
- `geom_col()` : apply column chart 
- `geom_bar()` : bar chart  
- `geom_boxplot()` : boxplot 
- `geom_hist()` : histogram

```{r}
# simple plot
ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() 

# layered plot

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") # here we are adding a linear model line to the plot


```

### Step 4 - Faceting the data

We can also split up our data by categorical variables in R. This is
called faceting. This is a very useful technique to use if we want to communicate differences between populations or timepoints (or other categorical variables). We have 3 options of faceting: \* With the aes choices (i.e., color, fill, shape, size). \* With `facet_wrap()` for faceting by 1 variable. \* With `facet_grid()` for faceting by 1 or more variables.

```{r}

# faceting by shape in aes
ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, 
           color = bill_length_mm, shape = species)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") # here we are adding a linear model line to the plot

# faceting with facet_wrap()

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") + # here we are adding a linear model line to the plot
  facet_wrap(~species)

# faceting with facet_grid()

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") + # here we are adding a linear model line to the plot
  facet_grid(~species ~sex)


```

### Step 5 - Changing the styling with them

If we want to change the visual display of the plot, we can use the `theme()` function, and use/combine this with pre-set themes. Here are some we can use:

-   theme_classic() - refined, classic, and does the job for APA formats
-   theme_bw()
-   theme_dark()
-   theme_minimal()

```{r warning = FALSE}
## notice that warning is set to FALSE in the header. This is to remove the warning message that has been created in the plot, and declutter for this workshop. 

# customising theme
ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") + # here we are adding a linear model line to the plot
  facet_wrap(~species) +
  theme(
    legend.position = "top" # change legend position inside of theme. 
    # you can try "top", "bottom", "right", "left", "none"
      ) 

# preset theme ontop of customised theme

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") + # here we are adding a linear model line to the plot
  facet_wrap(~species) +
  theme(
    legend.position = "top" # change legend position inside of theme. 
    # you can try "top", "bottom", "right", "left", "none"
      ) +
  theme_classic()
```

### Step 6 - Adding labels

Finally, we will make our secondary school science teachers proud - we will add labels to our plots. Jokes aside, labels are essential for communicating our plots with greater clarity. In R, it also gives us the opportunity to tidy up our variable names, so that they go from being computer friendly, to being human friendly.

```{r warning = FALSE}

ggplot(data = tidy_data,
       aes(x = flipper_length_mm, y = body_mass_g, color = bill_length_mm)
       ) + # we use the + to add layers to our ggplot.
  geom_point() +
  geom_smooth(method = "lm") + # here we are adding a linear model line to the plot
  facet_wrap(~species) +
  theme(
    legend.position = "top" # change legend position inside of theme. 
    # you can try "top", "bottom", "right", "left", "none"
      ) +
  theme_classic() + 
  labs(x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Bill Length (mm)",
       title = "Plot comparing Physiological measures of Penguin Species")

```

## Self Tasks

Time to use ggplot to visually explore the following questions:

### Task 1 - Does penguin flipper length differ by island?

```{r }

```

### Task 2 - To what extent does bill depth and sex predict bill length?

```{r}

```

### Task 3 - What is the distributuion of the different penguin species across the different islands?

```{r}

```

## End of Session

